The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2021-01-19 journal paper/content/related_work.tex
space saver

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11]{slivkins-MABbook}),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    (\eg see \citet{DynAuctions-survey10} for a survey),
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09,Transform-ec10-jacm},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13,Krause-www13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

@@

Several papers study competition between two principals who run algorithms but do not interact, directly or indirectly, until the very end of the game. \cite{Ufuk-jeea15} consider a ``research competition" between two firms racing towards a big discovery.
Each firm deploys a bandit algorithm with two arms, corresponding to safe and risky lines of research. The firms do not interact until one of them makes the discovery and wins the game. In the ``dueling algorithms" framework of \citet{DuelingAlgs-stoc11}, each principals runs an algorithm for the same problem. All inputs are observable at once, and principals' payoffs are binary (win/lose). \citet{ben2017best, ben2019regression} study competition between ``offline" machine learning algorithms. In comparison, we study a ``product competition" in which the two firms interact continuously (via the customers' choices), accrue rewards incrementally, and compete for individual customers.

A long line of work from electrical engineering and computer science, starting from
\citet{MultiPlayerMAB-Poor08,MultiPlayerMAB-Liu10,MultiPlayerMAB-Anima11}, focuses on competition for resources, not competition for consumers. Specifically, this literature targets an application to \emph{cognitive radios}, where multiple radios transmit simultaneously in a shared medium and compete for bandwidth. Each radio chooses channels over time using a multi-armed bandit algorithm. This work studies a repeated game between bandit algorithms, and focuses on designing algorithms which work well in this game, under various assumptions on communication, synchronization and collisions.

@@

The literature on learning-by-doing vs. competition \citep[\eg][]{gilbert1982preemptive, fudenberg1983learning, dasgupta1988learning, cabral1994learning}
studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its production costs. Our model differs in several important respects. First, firms learn to improve product quality rather than to reduce production costs. Second, the firms' current actions have consequences (via reputation and/or data collected by the algorithm) that directly impact consumer choices in the future. Third, we endogenize the technology behind learning-by-doing by explicitly considering bandit algorithms.

@@

\asedit{While we use first-mover advantage and agents' decision rule,} classic ``market competitiveness" measures, such as the Lerner Index or the Herfindahl-Hirschman Index
\citep{tirole1988theory}, are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
%\gadelete{Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. \footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.} This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.}
Second, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, mcfadden1974measurement, matvejka2015rational}.
Choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

>>>>

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11]{slivkins-MABbook}),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    (\eg see \citet{DynAuctions-survey10} for a survey),
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

@@

Several papers study competition between two principals who run algorithms but do not interact, directly or indirectly, until the very end of the game. \cite{Ufuk-jeea15} consider a ``research competition" between two firms racing towards a big discovery.
Each firm deploys a bandit algorithm with two arms, corresponding to safe and risky lines of research. The firms do not interact until one of them makes the discovery and wins the game. In the ``dueling algorithms" framework of \citet{DuelingAlgs-stoc11}, each principals runs an algorithm for the same problem. All inputs are observable at once, and principals' payoffs are binary (win/lose). \citet{ben2019regression} study competition between ``offline" machine learning algorithms. In comparison, we study a ``product competition" in which the two firms interact continuously (via the customers' choices), accrue rewards incrementally, and compete for individual customers.

%A long line of work from electrical engineering and computer science, starting from
%\citet{MultiPlayerMAB-Poor08,MultiPlayerMAB-Liu10,MultiPlayerMAB-Anima11}, focuses on competition for resources, not competition for consumers. Specifically, this literature targets an application to \emph{cognitive radios}, where multiple radios transmit simultaneously in a shared medium and compete for bandwidth. Each radio chooses channels over time using a multi-armed bandit algorithm. This work studies a repeated game between bandit algorithms, and focuses on designing algorithms which work well in this game, under various assumptions on communication, synchronization and collisions.

@@

The literature on learning-by-doing vs. competition \citep[\eg][]{fudenberg1983learning, dasgupta1988learning, cabral1994learning}
studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its production costs. Our model differs in several important respects. First, firms learn to improve product quality rather than to reduce production costs. Second, the firms' current actions have consequences (via reputation and/or data collected by the algorithm) that directly impact consumer choices in the future. Third, we endogenize the technology behind learning-by-doing by explicitly considering bandit algorithms.

@@

%\asedit{While we use first-mover advantage and agents' decision rule,} classic ``market competitiveness" measures, such as the Lerner Index or the Herfindahl-Hirschman Index
%\citep{tirole1988theory}, are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
%\gadelete{Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. \footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.} This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.}
Second, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, matvejka2015rational}.
Choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

### 2021-01-19 journal paper/content/sec-bg.tex
space saver

The optimal regret rates are achieved by several adaptive-exploration  algorithms, of which the most known are
Thompson Sampling \citep{Thompson-1933,TS-survey-FTML18},%
\footnote{While Thompson Sampling dates back to 1933 and is probably the best-known bandit algorithm, its regret has not been understood until recently \citep{Shipra-colt12,Kaufmann-alt12,Shipra-aistats13}.}
UCB1 \citep{bandits-ucb1},
and Successive Elimination \citep{EvenDar-icml06}.%
\footnote{A substantial follow-up work on more ``refined" regret rates is not as relevant to this paper.}
These algorithms are very simple to describe. Focus on one round and consider the posterior distribution and/or the confidence interval on each arm's mean reward. Thompson Sampling draws a sample (``score") from each arm's posterior distribution, and picks an arm with the largest score. UCB1 picks an arm with the largest upper confidence bound. Successive Elimination eliminates an arm once it is worse than some other arm with high confidence, and chooses uniformly among the remaining arms.

@@

The greedy algorithm can perform well \emph{sometimes} in a more general model of \emph{contextual bandits}, where auxiliary payoff-relevant signals, a.k.a. contexts, are observed before each round. This phenomenon has been observed in practice
\citep{practicalCB-arxiv18}, and in theory \citep{kannan2018smoothed,bastani2017exploiting,externalities-colt18} under (very) substantial assumptions. The prevalent intuition is that the diversity of contexts can --- under some conditions and to a limited extent --- substitute for explicit exploration.

>>>>

The optimal regret rates are achieved by several adaptive-exploration  algorithms, of which the most known are
Thompson Sampling \citep{Thompson-1933,TS-survey-FTML18},%
\footnote{While Thompson Sampling dates back to 1933 and is probably the best-known bandit algorithm, its regret has not been understood until recently \citep{Shipra-colt12}.}
UCB1 \citep{bandits-ucb1},
and Successive Elimination \citep{EvenDar-icml06}.%
\footnote{A substantial follow-up work on more ``refined" regret rates is not as relevant to this paper.}
These algorithms are very simple to describe. Focus on one round and consider the posterior distribution and/or the confidence interval on each arm's mean reward. Thompson Sampling draws a sample (``score") from each arm's posterior distribution, and picks an arm with the largest score. UCB1 picks an arm with the largest upper confidence bound. Successive Elimination eliminates an arm once it is worse than some other arm with high confidence, and chooses uniformly among the remaining arms.

@@

The greedy algorithm can perform well \emph{sometimes} in a more general model of \emph{contextual bandits}, where auxiliary payoff-relevant signals, a.k.a. contexts, are observed before each round. This phenomenon has been observed in practice
\citep{practicalCB-arxiv18}, and in theory \citep{kannan2018smoothed,externalities-colt18} under (very) substantial assumptions. The prevalent intuition is that the diversity of contexts can --- under some conditions and to a limited extent --- substitute for explicit exploration.

### 2021-01-19 journal paper/ec19paper/content/inverted_u.tex
Some edits after Guy's pass, removing some color-edits.
space saver

We investigate the firms' market shares when they choose different algorithms (otherwise, by symmetry both firms get half of the agents). We report the market shares for \gaedit{each instance in} Table~\ref{fig:market_share}. We find that $\DG$ is a weakly dominant strategy for the Heavy-Tail and Uniform instances, as long as $T_0$ is sufficiently small. However, \Thompson is a weakly dominant strategy for the Needle-in-Haystack instance. We find that for a sufficiently small $T_0$, \DynamicGreedy yields more than half the market against \Thompson,  but achieves similar market share vs. \DynamicGreedy and \DynamicEpsGreedy. By our tie-breaking rule, (\DynamicGreedy,\DynamicGreedy) is the only pure-strategy equilibrium.

@@

We attribute the prevalence of \DynamicGreedy on exploration-disadvantaged MAB instances to its prevalence on the initial ``exploration disadvantage period", as described in Section~\ref{sec:isolation}. Increasing the warm start length $T_0$ makes this period shorter: indeed, considering the relative reputation trajectory in Figure~\ref{relative_rep_plots} (left), increasing $T_0$ effectively shifts the starting time point to the right. This is why it helps \DynamicGreedy if $T_0$ is small.

>>>>

We investigate the firms' market shares when they choose different algorithms (otherwise, by symmetry both firms get half of the agents). We report the market shares for each instance in Table~\ref{fig:market_share}. We find that $\DG$ is a weakly dominant strategy for the Heavy-Tail and Uniform instances, as long as $T_0$ is sufficiently small. However, \Thompson is a weakly dominant strategy for the Needle-in-Haystack instance. We find that for a sufficiently small $T_0$, \DynamicGreedy yields more than half the market against \Thompson,  but achieves similar market share vs. \DynamicGreedy and \DynamicEpsGreedy. By our tie-breaking rule, (\DynamicGreedy,\DynamicGreedy) is the only pure-strategy equilibrium.

@@

We attribute the prevalence of \DynamicGreedy on exploration-disadvantaged MAB instances to its prevalence on the initial ``exploration disadvantage period", as described in Section~\ref{sec:isolation}. Increasing the warm start length $T_0$ makes this period shorter: indeed, considering the relative reputation trajectory in Figure~\ref{relative_rep_plots} (left), increasing $T_0$ effectively shifts the starting time point to the right. This is why it helps \DynamicGreedy if $T_0$ is small.\footnote{\gaedit{In Appendix Section~\ref{sec:perf_in_iso_revisited} we further explore the reason why the mean reputation trajectories do not suffice to explain the results observed under competition. The primary conclusion of the exercise is that the distribution of rewards of the algorithm is important and in some scenarios the mean is not always a good measure of the typical value of the distribution of rewards. As a result, it is not always predictive of results under competition whereas the relative reputation statistic is more closely related to how agents decide between the principals and thus more predictive of outcomes in the competition game.}}

### 2021-01-19 journal paper/ec19paper/content/revisited.tex
add changes
space saver

\subsection{Performance in Isolation, Revisited}\label{sec:revisited}

We saw in Section~\ref{sec:competition} that mean reputation trajectories do not suffice to explain the outcomes under competition. Let us provide more evidence and intuition for this.

@@

However, we find a more nuanced picture. For example, in Figure \ref{fig:market_share} we see that $\DynamicGreedy$ attains a larger market share than $\DynamicEpsGreedy$ even for large warm starts. We find that this also holds for $K = 3$ arms and longer time horizons, see the supplement for more details. We conclude \asedit{that Conjecture~\ref{conj:mean-trajectories} is false:}

>>>>

We saw in Section~\ref{sec:competition} that mean reputation trajectories do not suffice to explain the outcomes under competition. In this section we provide more evidence and intuition for this.

@@

However, we find a more nuanced picture. For example, in Figure \ref{fig:market_share} we see that $\DynamicGreedy$ attains a larger market share than $\DynamicEpsGreedy$ even for large warm starts. We find that this also holds for $K = 3$ arms and longer time horizons, see the supplement for more details. We conclude that Conjecture~\ref{conj:mean-trajectories} is false:
